Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-21 results.tex
some changes to wpe, results sections

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images
in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
to the right of the dashed line.  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and deterministic, and if one
uses a prediction technique that is based a nonlinear model
(LMA)---rather than a method that simply predicts the running mean
(\naive) or one that uses a linear model (ARIMA)---the MASE score is
much better: $0.050 \pm 0.001$.  This prediction is 20 times more
accurate than a random-walk forecast, which is more in line with the
level of predictive structure that the low WPE value suggests is
present in the signal.

>>>>

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images
in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
to the right of the dashed line.  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and \alert{largely} deterministic, and if one
uses a prediction technique that is based a nonlinear model
(LMA)---rather than a method that simply predicts the running mean
(\naive) or one that uses a linear model (ARIMA)---the MASE score is
much better: $0.050 \pm 0.001$.  This prediction is 20 times more
accurate than a random-walk forecast, which is more in line with the
level of predictive structure that the low WPE value suggests is
present in the signal.

### 2014-04-21 wpe.tex
some changes to wpe, results sections

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems is, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
not useful in our context, where we want to measure the entropy to
quantify predictability.

@@

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE is dominated by the noise about
the fixed points driving it to $\approx 1$, which in some sense hides
the fact that the signal is actually quite structured.  In this
situation, the weighted permutation entropy measures on switching
rather than the stochastic fluctuations. To accomplish this, the
\emph{weight} of a permutation is taken into account:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.

>>>>

Previous approaches to measuring temporal complexity via the Shannon
entropy rate \cite{Shannon1951, mantegna1994linguistic} required
categorical data: $x_i \in \mathcal{S}$ for some finite or countably
infinite \emph{alphabet} $\mathcal{S}$.  Data taken from real-world
systems is, however, effectively\footnote{Measurements from
  finite-precision sensors are discrete, but data from modern
  high-resolution sensors are, for the purposes of entropy
  calculations, effectively continuous.}  real-valued.  To analyze
real-valued data using a method that requires categorical values, one
must discretize the data---typically by binning.  Unfortunately, this
is rarely a good solution, as the binning of the values introduces
spurious dynamics~\cite{bollt2001}.  The field of symbolic dynamics
offers discretization methods that do not disturb the intrinsic
behavior.  These methods are, however, fragile in the face of noise;
worse yet, they require knowledge of the underlying system.  This is
not useful in our context, where we want to measure the entropy to
quantify predictability of an arbitrary time-series.

@@

Here we will be utilizing a variation of the basic permutation entropy
technique, the \emph{weighted permutation entropy} (WPE), which was
introduced in~\cite{fadlallah2013}.  The intent behind the weighting
is to correct for observational noise that is larger than the trends
in the data, but smaller than the larger scale features.  Consider,
for example, a signal that switches between two fixed points and
contains some additive noise. The PE is dominated by the noise about
the fixed points driving it to $\approx 1$, which in some sense hides
the fact that the signal is actually quite structured.  In this
situation, the terms of the weighted permutation entropy are dominated
by the switching rather than the stochastic fluctuations. To 
accomplish this, the \emph{weight} of a permutation is taken into
account:
\begin{align*}
  w(x_{i+1}^\ell) = \frac{1}{\ell} \sum_{j = i}^{i+\ell}
                      \left( x_j - \bar{x}_{i+1}^\ell \right)^2
\end{align*}
where $x_{i+1}^\ell$ is a sequence of values $x_{i+1}, \ldots,
x_{i+\ell}$, and $\bar{x}_{i+1}^\ell$ is the arithmetic mean of
those values.
